Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-08-10 analysis/paper/paper.Rmd
Minor edits

# Introduction
Variation in population patterns are regarded as important drivers and potential consequences of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at their time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

@@

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf. @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to be older than around 9300 BCE in the region [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to the final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. A common approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve [e.g. @breivik2014], and sometimes adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use, and limits the further inferential steps that can be taken. Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022; @timpson2021], found by using the values of the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic [e.g. @bang-andersen2005; @fuglestvedt2018, 16--17; @lindblom1984; @bjerck2008, 103--105], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE. However, properly adjudicating this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

@@

The main results was that the RSPD is consistent with the logistic model fit to the data. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD. Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations. Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions representing potential lines for further interrogation.

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. This is also relevant for the treatment of RSPDs, which in in Norwegian archaeology have usually been treated as constantly and uniformly impacted by population density through the period. We suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a determining contributor to the absence of radiocarbon dates from this earliest period. The degree to which mobility patterns might have also and variably impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].

Some lines of inquiry in the future could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022]. Furthermore, aligning these different lines of evidence will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see e.g. @lundstrom2021; @manninen2023].

Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application. While we believe the approach represents an improvement over previous approaches, the results should therefore considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.

>>>>

# Introduction
Variation in population patterns are regarded as important drivers and potential consequences of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2023; see also @persson2009].

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at their time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

@@

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf. @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a directly coherent statistical procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments, maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and, it is worth underscoring, that deviations from these rejected null models do not in themselves give statistical justification of the interpretation of any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves -- statistical or otherwise.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to hvae an older date than around 9300 BCE in the region [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. One approach to shoreline dating has been to find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve [e.g. @breivik2014]. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error to a point-estimate, as mentioned above, does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022; @timpson2021], found by using the values of the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use to avoid over-fitting, and provides a measure for the relative performance of the models.

@@

For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values, of the compared models, the logistic model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the Mesolithic, contrasted by a general stationary frequency of ^14^C-dated material. Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic [e.g. @bang-andersen2005; @fuglestvedt2018, 16--17; @lindblom1984; @bjerck2008, 103--105], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

@@

The main results was that the RSPD is consistent with the logistic model fit to the data. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD. Some of the deviations from the model were then narratively explored to suggest some possible reasons for the reasons and nature of these deviations. Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. This is also relevant for the treatment of RSPDs, which in in Norwegian archaeology have usually been treated as constantly and uniformly impacted by population density through the period. We suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a determining contributor to the absence of radiocarbon dates from this earliest period. The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022; @lundstrom2023]. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018].

Seeing as the implemented method for shoreline dating was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties and limitations associated with its application. While we believe the approach represents an improvement over previous approaches, the results should therefore be considered a cautions first exploration of the implications of employing a probabilistic method for shoreline dating when evaluating large scale trends in the temporal frequency distribution of coastal sites in the region.
